---
# Document Setup ---------------------------------------------------------

# Set document details.
title: 'Playing Tougher for One’s Size: \linebreak Positional Physicality in the NHL'
author: 'Rento Saijo^1^, Sian Lee^2^, Jeff Moher^1^, Yeongjun Seo^3^\*'
affiliations: '^1^ Connecticut College; ^2^ Michigan State University; ^3^ University of Tennessee, Knoxville'
corresponding: '\* Corresponding author: yseo4@utk.edu'
date: ''

# Configure PDF output.
format:
  pdf:
    documentclass: article
    pdf-engine: pdflatex
    fontsize: 10pt
    papersize: letter
    geometry: margin=1in
    fontfamily: lmodern
    indent: true
    fig-width: 6.4
    fig-height: 2.4
    fig-align: center
    number-sections: false

    # Configure LaTeX formatting.
    include-in-header:
      text: |
        % Page Layout --------------------------------------------------------

        % Load packages.
        \usepackage{fancyhdr, lastpage, titling}

        % Configure equations and paragraphs.
        \allowdisplaybreaks
        \setlength{\parindent}{0pt}

        % Format title and author.
        \pretitle{\begin{center}\Large\scshape}
        \setlength{\droptitle}{-3em}
        \preauthor{\begin{center}\large\scshape\begin{tabular}[t]{c}}

        \postauthor{\end{tabular}\par\end{center}\vspace{-2em}}
        \predate{}
        \postdate{}

        % Configure page footer.
        \pagestyle{fancy}
        \fancyhf{}
        \fancyfoot[C]{\textsc{Page \thepage\ of \pageref*{LastPage}}}
        \renewcommand{\headrulewidth}{0pt}
  gfm:
    wrap: none

# Configure code output.
execute:
  echo: false
  message: false
  warning: false


---

```{r}
#| label: setup
#| include: false
# Abstract Setup ---------------------------------------------------------

# Read completed results and document metadata.
base::source('../../R/abstract.R')
abstract <- prepare_abstract(base::readRDS('../../data/analysis_data.rds')$a3z)
metadata <- rmarkdown::yaml_front_matter(knitr::current_input())

# Configure vector graphics with template typography.
knitr::opts_chunk$set(dev = 'cairo_pdf', dev.args = base::list(family = 'Latin Modern Roman', pointsize = 10))
```

```{=latex}
\begingroup
\centering
```

```{r}
#| label: affiliations
#| results: asis
# Display author affiliations and correspondence.
base::cat(metadata$affiliations, '\n\n', metadata$corresponding, '\n')
```

```{=latex}
\par
\endgroup

% Preserve first-page footer and space below affiliations.
\thispagestyle{fancy}
\par\addvspace{\baselineskip}
```

## Introduction

Adding size to an NHL roster does not guarantee greater physical presence. Contact, puck battles, and contested space shape how a player plays tough, yet height, weight, and hit totals say little about whether that presence exceeds what a player's frame suggests. We develop calibrated size above expected (CSAx) to capture this distinction and ask whether it agrees with earlier scouting descriptions and illuminates NHL roles, postseason behavior, and roster persistence.

## Methods

We combine NHL events with All Three Zones' manually tracked five-on-five play from 2021-22 through 2024-25, covering `r base::format(abstract$forwardSeasons, big.mark = ',')` forward-seasons and `r abstract$defenseSeasons` defenseman-seasons. Separate ridge regressions learn the body size implied by shared direct measures, including hits delivered and received, and position-specific indirect measures, including dump-in recoveries for forwards and entry denials for defensemen. We calibrate held-out predictions against listed height and weight and standardize the residual within position and season. Higher CSAx denotes more physical presence than expected for a player's frame. One rater, blinded to player identity and CSAx, codes mentions of checking, puck battles, and other active engagement in official NHL draft-year scouting reports published before the studied seasons. We examine special-teams use, playoff contact, and next-season continuation and ice time, accounting for listed size, age, usage, scoring, relative shot attempts, and season. Playoff comparisons use regular-season games excluded from CSAx construction and each team's first four playoff games, giving every qualifier the same window regardless of advancement.

## Results

Scouting descriptions of active engagement precede play and align with CSAx: players with such descriptions average 0.87 standardized units higher among forwards and 0.65 among defensemen (Figure 1). This independent agreement supports the intended interpretation.

<!-- abstract-display-start -->

```{=latex}
\par\addvspace{0.5\baselineskip}
```

```{r}
#| label: scouting
#| results: asis
# Show scouting observations or their textual equivalent.
if (knitr::is_latex_output()) {
  base::print(plot_abstract_scouting(abstract))
} else {
  base::cat(abstract$scoutingText, '\n\n')
}
```

```{=latex}
\par
```

**Figure 1.** Player-average CSAx by scouting description. Points represent players; diamonds mark group means. Differences show 95% confidence intervals. The high-CSAx defenseman without active engagement description may serve as an example of how our metric can evaluate "tough" players who are sometimes overlooked by scouts due to their relative size; we examine this case further in the full paper.

<!-- abstract-display-end -->

```{=latex}
\newpage
```

Higher-CSAx players in both positions spend a smaller share of their ice time on power plays and a larger share killing penalties (Table 1), linking the profile to distinct assignments. In the postseason, higher-CSAx forwards show smaller proportional increases in delivered hits than lower-CSAx forwards; the defenseman contrast remains uncertain. Finally, comparing CSAx +1 with -1, adjusted probabilities of reaching 300 NHL minutes next season rise by 2.92 percentage points for forwards and 4.23 for defensemen. Among those who return, however, minutes per game show no clear increase. The profile may therefore relate to roster persistence without implying expanded playing time.

<!-- abstract-display-start -->

```{=latex}
\par\addvspace{\baselineskip}
```

**Table 1.** Adjusted CSAx +1 versus -1 comparisons (two standard deviations apart; 95% player-clustered confidence intervals in parentheses). pp means percentage points; ice time covers next-season returners only. The hit-rate ratio compares playoff-to-baseline changes; below one means a smaller proportional increase.

```{r}
#| label: applications
#| results: asis
# Present stored associations on comparable score contrasts.
knitr::kable(abstract$applications, format = 'pipe', align = base::c('l', 'r', 'r'))
```

Linear deployment and returner-ice-time models and logistic continuation models adjust for listed size, age and age squared, current-season games dressed and ice time per game, five-on-five points per 60, relative shot attempts, and season. Playoff Poisson models include player-season effects, five-on-five ice-time offsets, and postseason interactions with CSAx and these controls.

<!-- abstract-display-end -->

## Conclusion

CSAx translates a familiar scouting judgment into a positional measure that complements listed size, hits, and scoring statistics. The scouting, deployment, and continuation patterns suggest a tool for finding physical profiles relevant to specialized roles and roster decisions. Selectively tracked games, one scouting rater, and weak defensive size prediction in 2024-25 limit how broadly we can generalize.
