---
# Document Setup ---------------------------------------------------------

# Set document details.
title: 'Playing Tougher for One’s Size: \linebreak Positional Physicality in the NHL'
author: 'Rento Saijo^1^, Sian Lee^2^, Yeongjun Seo^3^, Jeff Moher^1^'
affiliations: '^1^ Connecticut College; ^2^ Michigan State University; ^3^ University of Tennessee, Knoxville'
date: ''

# Configure PDF output.
format:
  pdf:
    documentclass: article
    pdf-engine: pdflatex
    fontsize: 10pt
    papersize: letter
    geometry: margin=1in
    fontfamily: lmodern
    indent: true
    fig-width: 6.4
    fig-height: 2.6
    fig-align: center
    number-sections: false

    # Configure LaTeX formatting.
    include-in-header:
      text: |
        % Page Layout --------------------------------------------------------

        % Load packages.
        \usepackage{fancyhdr, lastpage, titling}

        % Configure equations and paragraphs.
        \allowdisplaybreaks
        \setlength{\parindent}{0pt}

        % Format title and author.
        \pretitle{\begin{center}\Large\scshape}
        \setlength{\droptitle}{-3em}
        \preauthor{\begin{center}\large\scshape\begin{tabular}[t]{c}}

        \postauthor{\end{tabular}\par\end{center}\vspace{-2em}}
        \predate{}
        \postdate{}

        % Configure page footer.
        \pagestyle{fancy}
        \fancyhf{}
        \fancyfoot[C]{\textsc{Page \thepage\ of \pageref*{LastPage}}}
        \renewcommand{\headrulewidth}{0pt}
  gfm:
    wrap: none

# Configure code output.
execute:
  echo: false
  message: false
  warning: false


---

```{r}
#| label: setup
#| include: false
# Abstract Setup ---------------------------------------------------------

# Read completed results and document metadata.
base::source('../../R/abstract.R')
abstract <- prepare_abstract(base::readRDS('../../data/analysis_data.rds')$a3z)
metadata <- rmarkdown::yaml_front_matter(knitr::current_input())

# Configure vector graphics with template typography.
knitr::opts_chunk$set(dev = 'cairo_pdf', dev.args = base::list(family = 'Latin Modern Roman', pointsize = 10))
```

```{=latex}
\begingroup
\centering
```

```{r}
#| label: affiliations
#| results: asis
# Display author affiliations.
base::cat(metadata$affiliations, '\n')
```

```{=latex}
\par
\endgroup

% Preserve first-page footer and space below affiliations.
\thispagestyle{fancy}
\par\addvspace{\baselineskip}
```

## Introduction

Adding size to a roster does not guarantee the physical presence a team needs. That presence emerges through contact, puck battles, and contested space. We develop calibrated size above expected (CSAx) to quantify playing tougher for one's size and examine what that profile means for NHL roles and continued participation.

## Methods

We combine NHL events with All Three Zones' manually tracked puck plays from 2021-22 through 2024-25, covering `r base::format(abstract$forwardSeasons, big.mark = ',')` forward-seasons and `r abstract$defenseSeasons` defenseman-seasons. Separate positional ridge regressions estimate the body size implied by direct contact and position-specific indirect physicality. We calibrate held-out predictions against listed height and weight, then standardize residuals within position and season. Higher CSAx indicates more physical presence than expected for that frame. Playoff comparisons use regular-season baselines excluded from score construction.

## Results

Earlier scouting reports offer independent support: players described as actively physical subsequently receive higher CSAx in both positions (Figure 1).

```{=latex}
\par\addvspace{\baselineskip}
```

```{r}
#| label: scouting
#| results: asis
# Show scouting observations or their textual equivalent.
if (knitr::is_latex_output()) {
  base::print(plot_abstract_scouting(abstract))
} else {
  base::cat(abstract$scoutingText, '\n\n')
}
```

```{=latex}
\par\addvspace{0.5\baselineskip}
```

**Figure 1.** Player-average CSAx by scouting description. Points represent players; diamonds mark means. Differences show 95% intervals. No mention does not establish soft play.

```{=latex}
\newpage
```

Across the full sample, higher-scoring players spend less of their ice time on power plays and more killing penalties (Table 1). When play shifts to the postseason, higher-scoring forwards show smaller proportional increases in delivered hits over each team's first four playoff games; defensive differences remain uncertain. Looking ahead, the adjusted probability of reaching 300 NHL minutes next season is 2.92 percentage points higher for forwards and 4.23 for defensemen at CSAx +1 versus -1. Among returners, however, minutes per game show no clear increase.

```{=latex}
\par\addvspace{\baselineskip}
```

**Table 1.** Adjusted CSAx +1 versus -1 comparisons (two standard deviations apart; 95% intervals in parentheses). pp means percentage points; ice time is next-season minutes per game among returners. The hit-rate ratio compares playoff-to-baseline rate changes; below one means a smaller proportional increase.

```{r}
#| label: applications
#| results: asis
# Present stored associations on comparable score contrasts.
knitr::kable(abstract$applications, format = 'pipe', align = base::c('l', 'r', 'r'))
```

## Conclusion

CSAx connects a familiar scouting judgment to measurable differences in deployment, playoff behavior, and NHL participation. Selective tracking, one scouting rater, and weak 2024-25 defensive size prediction limit generalization. For teams, CSAx broadens physicality evaluation beyond size and hit counts, helping identify players who bring more physical presence than their frame alone would suggest.
