---
# Document Setup ---------------------------------------------------------

# Set document details.
title: 'Playing Tougher for One’s Size: \linebreak Calibrated Size Above Expected in the NHL'
author: 'Rento Saijo^1^, Sian Lee^2^, Jeff Moher^1^, Yeongjun Seo^3^\*'
affiliations: '^1^ Connecticut College; ^2^ Michigan State University; ^3^ University of Tennessee, Knoxville'
corresponding_author: '\* Corresponding author: yseo4@utk.edu'
date: ''

# Configure PDF output.
format:
  pdf:
    documentclass: article
    pdf-engine: pdflatex
    fontsize: 10pt
    papersize: letter
    geometry: margin=1in
    fontfamily: lmodern
    indent: true
    fig-width: 6.4
    fig-height: 2.4
    fig-align: center
    number-sections: false

    # Configure LaTeX formatting.
    include-in-header:
      text: |
        % Page Layout --------------------------------------------------------

        % Load packages.
        \usepackage{fancyhdr, lastpage, titling}

        % Configure equations and paragraphs.
        \allowdisplaybreaks
        \setlength{\parindent}{0pt}

        % Format title and author.
        \pretitle{\begin{center}\Large\scshape}
        \setlength{\droptitle}{-3em}
        \preauthor{\begin{center}\large\scshape\begin{tabular}[t]{c}}

        \postauthor{\end{tabular}\par\end{center}\vspace{-2em}}
        \predate{}
        \postdate{}

        % Configure page footer.
        \pagestyle{fancy}
        \fancyhf{}
        \fancyfoot[C]{\textsc{Page \thepage\ of \pageref*{LastPage}}}
        \renewcommand{\headrulewidth}{0pt}
  gfm:
    wrap: none

# Configure code output.
execute:
  echo: false
  message: false
  warning: false


---

```{r}
#| label: setup
#| include: false
# Abstract Setup ---------------------------------------------------------

# Read completed results and document metadata.
base::source('../../R/abstract.R')
abstract <- prepare_abstract(base::readRDS('../../data/analysis_data.rds')$a3z)
metadata <- rmarkdown::yaml_front_matter(knitr::current_input())

# Configure graphics for each output format.
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(dev = 'cairo_pdf', dev.args = base::list(family = 'Latin Modern Roman', pointsize = 10))
} else {
  knitr::opts_chunk$set(dev = 'png', dpi = 300)
}
```

```{=latex}
\begingroup
\centering
```

```{r}
#| label: affiliations
#| results: asis
# Display author affiliations.
base::cat(metadata$affiliations, '\n\n', metadata$corresponding_author, '\n')
```

```{=latex}
\par
\endgroup

% Preserve first-page footer and space below affiliations.
\thispagestyle{fancy}
\par\addvspace{\baselineskip}
```

## Introduction

In hockey, physical play involves contact and battles for the puck and space, but body size and hit counts alone do not quantify whether a player’s physical play exceeds or falls below expectations for their size. We develop calibrated size above expected (CSAx), a position-specific measure of physical play relative to listed size, to operationalize *playing tougher for one’s size*. We examine its agreement with scouting reports and links to special-teams deployment, postseason hitting, and next-season continuation.

## Methods

NHL event data and All Three Zones’ manually tracked five-on-five data were combined, covering 1,485 forward- and 829 defenseman-seasons from 2021–22 through 2024–25. Separate ridge regressions for forwards and defensemen estimate behavior-implied size by predicting standardized size derived from listed height and weight. Predictors include six shared direct measures (e.g., hits delivered and received), six forward indirect measures (e.g., dump-in recoveries), and three defenseman indirect measures (e.g., entry-denial share). Held-out predictions are calibrated against listed size, and residuals are standardized within position and season:

$$
\mathrm{CSAx}=\frac{r-\bar r}{s_r},\qquad r=\widehat S-\widehat m(S).
$$

Here, $S$ is standardized listed size, $\widehat S$ is behavior-implied size, $\widehat m(S)$ is the expected behavior-implied size at $S$, $r$ is the calibrated residual, and $\bar r$ and $s_r$ are the mean and standard deviation of these residuals within each position and season. Higher CSAx indicates greater behavior-implied size after accounting for listed size. One rater, blinded to identity and CSAx, codes mentions of physical traits (e.g., checking, battles) in official NHL draft-year scouting reports published before the studied seasons. Linear models assess special-teams shares and returner ice time; logistic models assess next-season continuation (at least 300 NHL minutes); and Poisson models compare hit rates within player-seasons with ice-time offsets. Analyses adjust for listed size, age and age squared, current-season games dressed and ice time per game, five-on-five points per 60, relative shot attempts, and season. We compare regular-season games excluded from CSAx construction with each team’s first four postseason games, a common window regardless of advancement.

## Results

Players with physical-trait mentions in earlier scouting reports have higher mean CSAx than those without, by 0.87 for forwards and 0.65 for defensemen (Figure 1).

<!-- abstract-display-start -->

```{r}
#| label: scouting-page-break
#| results: asis
# Place scouting figure at top of second PDF page.
if (knitr::is_latex_output()) base::cat('\\newpage\n')
```

```{r}
#| label: scouting
#| results: asis
# Show scouting observations.
base::print(plot_abstract_scouting(abstract))
```

```{=latex}
\par\addvspace{0.5\baselineskip}
```

**Figure 1.** Player-average CSAx by whether an earlier scouting report mentions active physical traits. Points represent players; diamonds mark group means. Mean differences, with 95% confidence intervals, suggest general agreement with earlier scouting descriptions.

<!-- abstract-display-end -->

```{=latex}
\par\addvspace{0.5\baselineskip}
```

Higher CSAx is associated with a smaller share of players’ total ice time on power plays and a larger share killing penalties in both positions (Table 1). Among forwards, higher CSAx is associated with a lower postseason-to-regular-season hit-rate ratio; the defensemen estimate is uncertain. Higher CSAx also predicts next-season continuation, but not ice time among returners.

<!-- abstract-display-start -->

```{r}
#| label: applications-page-break
#| results: asis
# Keep the application table together in the Word copy.
if (!knitr::is_latex_output()) base::cat('\\newpage\n')
```

```{=latex}
\par\addvspace{\baselineskip}
```

**Table 1.** Adjusted comparisons of CSAx +1 versus -1 (95% player-clustered confidence intervals in parentheses). Postseason estimates are ratios of postseason-to-baseline hit-rate ratios (+1 divided by -1); all other estimates are differences (+1 minus -1).

```{r}
#| label: applications
#| results: asis
# Present stored associations on comparable score contrasts.
knitr::kable(abstract$applications, format = 'pipe', align = base::c('l', 'r', 'r'))
```

<!-- abstract-display-end -->

## Conclusion

CSAx quantifies physical play relative to listed size, measuring *playing tougher for one’s size*. The high-CSAx defenseman without a physical-trait mention highlights a scouting application: identifying smaller players whose engagement may be overlooked by absolute size. Agreement with scouting and associations with penalty-kill use and next-season continuation support relevance to specialized roles. Continuation without greater ice time among returners suggests CSAx captures playing style relevant to NHL participation. Alongside hits and scoring, CSAx adds a size-relative perspective on toughness and roster fit.
